---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geographiclib

<!-- badges: start -->
<!-- badges: end -->

The goal of geographiclib is to wrap the awesome [GeographicLib](https://geographiclib.sourceforge.io/) library. 

Currently only reverse and forward for a simple MGRS conversion is exposed, WIP. 

Notes for work in progress: 

* no changes made to source files, check complains about kissfft.hh and src/GeographicLib/Makefile
* source code for *this* package is named "000_stuff_geographiclib.cpp" to differentiate from the original sources, which are all included
* see data-raw/GeographicLib.R for the obtaining source (at 2025-06-16 https://github.com/geographiclib/geographiclib 0a6067b74d2c5316afceb61e3a7a2b2f262960d8)

## Have questions? 

Other packages that include source from GeographicLib are these, none were suitable for my purposes: 

* geosphere for miscellaneous (I considered extending geosphere but it's not a very responsive project)
* nngeo for near neighbours
* geodist for distance calcs
* sf for distance calcs
* terra for various
* googlePolylines
* BH
* lwgeom
* s2
* mgrs (awesome, but not vectorized and uses the old military code)


## Installation

You can install the development version of geographiclib like so:

``` r
remotes::install_github("hypertidy/geographiclib")
```

## Example

This is a conversion for MGRS, please note this is really in testing stage and the details will change. 

```{r example}
library(geographiclib)

(code <- mgrs_fwd(c(147.325, -42.881)))
mgrs_rev(code)
```

The foward mode is vectorized on coordinate and precision value. 

```{r vectorized-mgrs}
pts <- cbind(runif(6, -180, 180), runif(6, -90, 90))
dput(pts)
mgrs_fwd(pts, precision = 0:5)
mgrs_fwd(pts, precision = 5:0)

```

Also it's fast. 

```{r speed, eval=FALSE}
x <- do.call(cbind, maps::world.cities[c("long", "lat")])
dim(x)
#[1] 43645     2
system.time(codes <- mgrs_fwd(x))
#   user  system elapsed 
#   0.04    0.00    0.04 
sample(codes, 10)
# [1] "37NCG3952467839" "31PBK7766746791" "36SWD3827984213" "35ULP9426067305" "45VUC7504263576"
# [6] "36RXV9463390163" "31UFT6135362533" "11SLT9551050534" "32TQQ0915128552" "32PMT1934289062"
sum(nchar(codes))
#[1] 654675
```   


## Code of Conduct
  
Please note that the geographiclib project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
