---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geographiclib

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/geographiclib/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/geographiclib/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

R interface to [GeographicLib](https://geographiclib.sourceforge.io/) for precise geodetic calculations on the WGS84 ellipsoid.

## Features

* **MGRS** - Military Grid Reference System encoding/decoding
* **UTM/UPS** - Universal Transverse Mercator and Universal Polar Stereographic projections
* **Transverse Mercator** - Custom TM projections with user-defined parameters
* **Lambert Conformal Conic** - Configurable LCC projection
* **Albers Equal Area** - Equal-area conic projection
* **Polar Stereographic** - Conformal polar projection with configurable scale
* **Azimuthal Equidistant** - Projection centered on any point
* **Cassini-Soldner** - Historical transverse cylindrical projection
* **Gnomonic** - Geodesics appear as straight lines
* **OSGB** - Ordnance Survey National Grid (Great Britain)
* **Geohash** - Geohash encoding/decoding with precision control
* **GARS** - Global Area Reference System (military)
* **Georef** - World Geographic Reference System (aviation)
* **Geodesic calculations** - Distance, bearing, paths on the ellipsoid (exact and fast)
* **Rhumb lines** - Constant-bearing paths (loxodromes)
* **Polygon area** - Accurate area/perimeter on the ellipsoid
* **Geocentric** - ECEF (Earth-Centered Earth-Fixed) coordinates
* **Local Cartesian** - East-North-Up (ENU) local coordinates
* **Ellipsoid** - WGS84 parameters, curvature, auxiliary latitudes

All functions are fully vectorized with rich metadata output.

## Installation

```r
remotes::install_github("hypertidy/geographiclib")
```

## Quick example

```{r example}
library(geographiclib)

# Coordinates to MGRS
pts <- cbind(lon = c(147.32, -74.01, 0.13),
             lat = c(-42.88, 40.71, 51.51))
(codes <- mgrs_fwd(pts))

# MGRS back to coordinates with full metadata
mgrs_rev(codes)
```

## Documentation

See the [package overview vignette](https://hypertidy.github.io/geographiclib/articles/geographiclib-overview.html) for a comprehensive tour of all features including:

- MGRS and Geohash encoding
- UTM/UPS and Lambert Conformal Conic projections
- Geodesic distance, bearing, and path calculations
- Polygon area on the ellipsoid

Full documentation at <https://hypertidy.github.io/geographiclib/>

## Performance

Fast C++ implementation - process tens of thousands of coordinates in milliseconds:

```{r speed, eval=FALSE}
x <- do.call(cbind, maps::world.cities[c("long", "lat")])
system.time(mgrs_fwd(x))       # 43,645 points
#>  user  system elapsed 
#>  0.04    0.00    0.04
```

## Code of Conduct

Please note that the geographiclib project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
