---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geographiclib

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/geographiclib/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/geographiclib/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of geographiclib is to wrap the awesome [GeographicLib](https://geographiclib.sourceforge.io/) library for precise geodetic calculations in R.

## Features

* **Fully vectorized MGRS conversion** - Convert thousands of coordinates in milliseconds
* **Rich reverse conversion output** - Get UTM/UPS coordinates, zones, convergence, scale factors, grid designators, and EPSG codes
* **Variable precision support** - Generate MGRS codes from 100km to 1m precision
* **Polar region support** - Handles UPS (Universal Polar Stereographic) zones automatically
* **Fast C++ implementation** - Built on GeographicLib's battle-tested geodetic algorithms

Currently MGRS (Military Grid Reference System) conversion is exposed. More GeographicLib functionality coming soon!

## Installation

You can install the development version of geographiclib like so:

``` r
remotes::install_github("hypertidy/geographiclib")
```

## Example

Convert coordinates to MGRS and back:

```{r example}
library(geographiclib)

(code <- mgrs_fwd(c(147.325, -42.881)))
mgrs_rev(code)
```

### Vectorized forward conversion

The `mgrs_fwd()` function is fully vectorized on both coordinates and precision:

```{r vectorized-mgrs}
pts <- cbind(runif(6, -180, 180), runif(6, -90, 90))
dput(pts)

# Variable precision: from 100km (0) to 1m (5)
mgrs_fwd(pts, precision = 0:5)

# Different precisions for each point
(code <- mgrs_fwd(pts, precision = 5:0))
```

### Rich reverse conversion output

The `mgrs_rev()` function returns a comprehensive data frame with 12 columns:

- **Geographic coordinates** (`lon`/`lat` in decimal degrees)
- **Projected coordinates** (`x`/`y` in meters, UTM/UPS)
- **Zone information** (`zone` number, `northp` hemisphere, `grid_zone` designator, `square_100km` ID)
- **Geodetic properties** (`convergence` angle in degrees, `scale` factor)
- **Precision level** (0-5, decoded from MGRS string)
- **CRS identification** (`crs` as EPSG codes for direct use with spatial packages)


```{r vectorized-rev}
mgrs_rev(code)
```

The reverse conversion returns the center point of each MGRS grid cell.

### Polar regions

Polar coordinates automatically use UPS (Universal Polar Stereographic):

```{r polar}
# North and South pole regions
polar_pts <- cbind(c(147, 148, -100), c(88, -88, -85))
polar_codes <- mgrs_fwd(polar_pts)
mgrs_rev(polar_codes)
```

Note that `zone = 0` indicates UPS projection, with dedicated EPSG codes (32661 for North, 32761 for South).

### Performance

It's fast - process tens of thousands of coordinates in milliseconds:

```{r speed, eval=FALSE}
# World cities dataset
x <- do.call(cbind, maps::world.cities[c("long", "lat")])
dim(x)
#[1] 43645     2

system.time(codes <- mgrs_fwd(x))
#   user  system elapsed 
#   0.04    0.00    0.04 

sample(codes, 10)
# [1] "37NCG3952467839" "31PBK7766746791" "36SWD3827984213" "35ULP9426067305" "45VUC7504263576"
# [6] "36RXV9463390163" "31UFT6135362533" "11SLT9551050534" "32TQQ0915128552" "32PMT1934289062"

sum(nchar(codes))
#[1] 654675
```

## Comparison with other packages

Several R packages include GeographicLib source code, but none provided the vectorized MGRS functionality needed:

* **mgrs** - MGRS support but not vectorized, uses older GEOTRANS code
* **geosphere** - Miscellaneous geodetic functions
* **sf, terra, s2** - Distance and geodetic calculations
* **geodist** - Fast distance calculations
* **nngeo** - Nearest neighbor operations
* **googlePolylines, BH, lwgeom** - Other geodetic utilities

## Development notes

* No changes made to GeographicLib source files
* Package source code named "000_*_geographiclib.cpp" to differentiate from original sources
* See `data-raw/GeographicLib.R` for source provenance (obtained 2025-06-16 from https://github.com/geographiclib/geographiclib commit 0a6067b)
* Note: R CMD check complains about kissfft.hh and src/GeographicLib/Makefile (non-package files from upstream)

## Code of Conduct
  
Please note that the geographiclib project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
