---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geographiclib

<!-- badges: start -->
<!-- badges: end -->

The goal of geographiclib is to wrap the awesome [GeographicLib](https://geographiclib.sourceforge.io/) library. 

Currently only reverse and forward for a simple MGRS conversion is exposed, WIP. 

Notes for work in progress: 

* no changes made to source files, check complains about kissfft.hh and src/GeographicLib/Makefile
* source code for *this* package is named "000_stuff_geographiclib.cpp" to differentiate from the original sources, which are all included
* see data-raw/GeographicLib.R for the obtaining source (at 2025-06-16 https://github.com/geographiclib/geographiclib 0a6067b74d2c5316afceb61e3a7a2b2f262960d8)

## Installation

You can install the development version of geographiclib like so:

``` r
remotes::install_github("hypertidy/geographiclib")
```

## Example

This is a conversion for MGRS, please note this is really in testing stage and the details will change. 

```{r example}
library(geographiclib)

(code <- mgrs_fwd(c(147.325, -42.881)))
mgrs_rev(code)
```

## Code of Conduct
  
Please note that the geographiclib project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
