---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geographiclib

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/geographiclib/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/geographiclib/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of geographiclib is to wrap the awesome [GeographicLib](https://geographiclib.sourceforge.io/) library for precise geodetic calculations in R.

## Features

* **Fully vectorized MGRS conversion** - Convert thousands of coordinates in milliseconds
* **Fully vectorized UTM/UPS conversion** - Direct access to Universal Transverse Mercator and Universal Polar Stereographic projections
* **Geodesic polygon area** - Accurate area and perimeter calculations on the WGS84 ellipsoid
* **Rich output** - Get projected coordinates, zones, convergence, scale factors, grid designators, and EPSG codes
* **Variable precision support** - Generate MGRS codes from 100km to 1m precision
* **Polar region support** - Handles UPS (Universal Polar Stereographic) zones automatically
* **Fast C++ implementation** - Built on GeographicLib's battle-tested geodetic algorithms

## Installation

You can install the development version of geographiclib like so:

``` r
remotes::install_github("hypertidy/geographiclib")
```

## MGRS - Military Grid Reference System

Convert coordinates to MGRS and back:

```{r example}
library(geographiclib)

(code <- mgrs_fwd(c(147.325, -42.881)))
mgrs_rev(code)
```

### Vectorized forward conversion

The `mgrs_fwd()` function is fully vectorized on both coordinates and precision:

```{r vectorized-mgrs}
pts <- cbind(runif(6, -180, 180), runif(6, -90, 90))
dput(pts)

# Variable precision: from 100km (0) to 1m (5)
mgrs_fwd(pts, precision = 0:5)

# Different precisions for each point
(code <- mgrs_fwd(pts, precision = 5:0))
```

### Rich reverse conversion output

The `mgrs_rev()` function returns a comprehensive data frame with 12 columns:

- **Geographic coordinates** (`lon`/`lat` in decimal degrees)
- **Projected coordinates** (`x`/`y` in meters, UTM/UPS)
- **Zone information** (`zone` number, `northp` hemisphere, `grid_zone` designator, `square_100km` ID)
- **Geodetic properties** (`convergence` angle in degrees, `scale` factor)
- **Precision level** (0-5, decoded from MGRS string)
- **CRS identification** (`crs` as EPSG codes for direct use with spatial packages)

```{r vectorized-rev}
mgrs_rev(code)
```

The reverse conversion returns the center point of each MGRS grid cell.

## UTM/UPS - Universal Transverse Mercator / Universal Polar Stereographic

Direct access to UTM/UPS projections for precise coordinate conversion:

```{r utmups-example}
# Forward: Geographic to UTM/UPS
pts <- cbind(lon = c(147.325, -63.22, 0, 45.67),
             lat = c(-42.881, 17.62, 88, 39.84))
(utm <- utmups_fwd(pts))
```

The forward conversion automatically selects the appropriate UTM zone (or UPS for polar regions) and returns:

- Projected coordinates (x/y in meters)
- Zone and hemisphere information
- Meridian convergence and scale factor
- EPSG CRS codes

```{r utmups-reverse}
# Reverse: UTM/UPS to Geographic
utmups_rev(utm$x, utm$y, utm$zone, utm$northp)
```

## Polygon Area - Geodesic area and perimeter

Compute accurate polygon area and perimeter on the WGS84 ellipsoid:

```{r polygon-area}
# Triangle: London - New York - Rio de Janeiro
pts <- cbind(
  lon = c(0, -74, -43),
  lat = c(52, 41, -23)
)
polygon_area(pts)
```

The area is returned in square meters and the perimeter in meters. The area is signed: positive for counter-clockwise polygons, negative for clockwise.

```{r polygon-area-km}
# Area in square kilometers
result <- polygon_area(pts)
abs(result$area) / 1e6
```

### Multiple polygons

Use the `id` argument to compute area for multiple polygons at once:

```{r polygon-multiple}
pts <- cbind(
  lon = c(0, -74, -43, 100, 110, 105, 120),
  lat = c(52, 41, -23, 10, 10, 20, 15)
)
polygon_area(pts, id = c(1, 1, 1, 2, 2, 2, 2))
```

### Polyline length

Set `polyline = TRUE` to compute the length of a path instead of a closed polygon:

```{r polyline}
# Great circle route length
route <- cbind(
  lon = c(151.2, -122.4, -0.1),  # Sydney - San Francisco - London

  lat = c(-33.9, 37.8, 51.5)
)
polygon_area(route, polyline = TRUE)
```

### Polar regions

Both MGRS and UTM/UPS automatically handle polar regions using UPS:

```{r polar}
# North and South pole regions
polar_pts <- cbind(c(147, 148, -100), c(88, -88, -85))

# MGRS in polar regions
polar_mgrs <- mgrs_fwd(polar_pts)
mgrs_rev(polar_mgrs)

# Direct UTM/UPS conversion
utmups_fwd(polar_pts)
```

Note that `zone = 0` indicates UPS projection, with dedicated EPSG codes (32661 for North, 32761 for South).

### Performance

It's fast - process tens of thousands of coordinates in milliseconds:

```{r speed, eval=FALSE}
# World cities dataset
x <- do.call(cbind, maps::world.cities[c("long", "lat")])
dim(x)
#[1] 43645     2

# MGRS conversion
system.time(codes <- mgrs_fwd(x))
#   user  system elapsed 
#   0.04    0.00    0.04 

# UTM/UPS conversion
system.time(utm <- utmups_fwd(x))
#   user  system elapsed 
#   0.03    0.00    0.03

sample(codes, 10)
# [1] "37NCG3952467839" "31PBK7766746791" "36SWD3827984213" "35ULP9426067305" "45VUC7504263576"
# [6] "36RXV9463390163" "31UFT6135362533" "11SLT9551050534" "32TQQ0915128552" "32PMT1934289062"

sum(nchar(codes))
#[1] 654675
```

## Comparison with other packages

Several R packages include GeographicLib source code, but none provided the vectorized MGRS, UTM/UPS, and polygon area functionality needed:

* **mgrs** - MGRS support but not vectorized, uses older GEOTRANS code
* **geosphere** - Miscellaneous geodetic functions including polygon area (spherical approximation)
* **sf, terra, s2** - Distance and geodetic calculations
* **geodist** - Fast distance calculations
* **nngeo** - Nearest neighbor operations
* **googlePolylines, BH, lwgeom** - Other geodetic utilities

## Development notes

* No changes made to GeographicLib source files
* Package source code named "000_*_geographiclib.cpp" to differentiate from original sources
* See `data-raw/GeographicLib.R` for source provenance (obtained 2025-06-16 from https://github.com/geographiclib/geographiclib commit 0a6067b)
* Note: R CMD check complains about kissfft.hh and src/GeographicLib/Makefile (non-package files from upstream)

## Code of Conduct
  
Please note that the geographiclib project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
