---
title: "Introduction to geographiclib"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to geographiclib}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(geographiclib)
```

The **geographiclib** package provides R bindings to the 
[GeographicLib](https://geographiclib.sourceforge.io/) C++ library for precise 
geodetic calculations. All functions are fully vectorized and return rich 
metadata.
 
## MGRS - Military Grid Reference System

Convert geographic coordinates to MGRS codes and back:

```{r mgrs}
# Forward conversion
(code <- mgrs_fwd(c(147.325, -42.881)))

# Reverse returns rich metadata
mgrs_rev(code)
```

Variable precision from 100km (0) to 1m (5):

```{r mgrs-precision}
mgrs_fwd(c(147.325, -42.881), precision = 0:5)
```

## UTM/UPS - Universal Transverse Mercator

Direct access to UTM projections with automatic zone selection:

```{r utmups}
pts <- cbind(lon = c(147.325, -74.006, 0), lat = c(-42.881, 40.7128, 88))
utmups_fwd(pts)
```

Reverse conversion requires zone and hemisphere:

```{r utmups-rev}
utm <- utmups_fwd(c(147.325, -42.881))
utmups_rev(utm$x, utm$y, utm$zone, utm$northp)
```

## Geohash

Encode locations as compact strings with hierarchical precision:

```{r geohash}
# Default length 12 (~19mm precision)
(gh <- geohash_fwd(c(147.325, -42.881)))

# Truncation preserves containment
substr(gh, 1, c(4, 6, 8))

# Reverse shows resolution
geohash_rev(gh)
```

Resolution table for different lengths:

```{r geohash-res}
geohash_resolution(c(4, 6, 8, 10, 12))
```

## Lambert Conformal Conic Projection

The LCC projection is widely used for aeronautical charts and regional coordinate systems:

```{r lcc}
# Single standard parallel (tangent cone)
pts <- cbind(lon = c(-100, -99, -98), lat = c(40, 41, 42))
lcc_fwd(pts, lon0 = -100, stdlat = 40)

# Two standard parallels (secant cone)
lcc_fwd(pts, lon0 = -96, stdlat1 = 33, stdlat2 = 45)
```

## Geodesic Calculations

Solve geodesic problems on the WGS84 ellipsoid.

### Inverse problem: distance and bearing between points

```{r geodesic-inverse}
# London to New York
geodesic_inverse(c(-0.1, 51.5), c(-74, 40.7))
```

### Direct problem: destination from start, bearing, and distance

```{r geodesic-direct}
# 1000km east from London
geodesic_direct(c(-0.1, 51.5), azi = 90, s = 1000000)
```

### Geodesic paths

Generate points along the shortest path between two locations:

```{r geodesic-path}
path <- geodesic_path(c(-0.1, 51.5), c(-74, 40.7), n = 5)
path
```

### Distance matrix

```{r geodesic-matrix}
cities <- cbind(
  lon = c(-0.1, -74, 139.7, 151.2),
  lat = c(51.5, 40.7, 35.7, -33.9)
)
rownames(cities) <- c("London", "NYC", "Tokyo", "Sydney")

# Distance matrix in km
round(geodesic_distance_matrix(cities) / 1000)
```

## Rhumb Lines (Loxodromes)

Rhumb lines are paths of constant bearing. They are not the shortest path 
(geodesics are), but they maintain a constant compass heading - useful for 
navigation.

```{r rhumb}
# Rhumb distance vs geodesic: London to New York
geodesic_inverse(c(-0.1, 51.5), c(-74, 40.7))$s12 / 1000  # km
rhumb_inverse(c(-0.1, 51.5), c(-74, 40.7))$s12 / 1000     # km (longer!)

# Generate rhumb line path
rhumb_path(c(-0.1, 51.5), c(-74, 40.7), n = 5)
```

Key properties: east-west rhumb lines maintain constant latitude, north-south 
rhumb lines maintain constant longitude:

```{r rhumb-ew}
# Heading due east maintains latitude
rhumb_direct(c(0, 45), azi = 90, s = 1000000)
```

## Polygon Area

Compute accurate polygon area and perimeter on the ellipsoid:
 
```{r polygon-area}
# Triangle: London - New York - Sydney
triangle <- cbind(
  lon = c(-0.1, -74, 151.2),
  lat = c(51.5, 40.7, -33.9)
)
result <- polygon_area(triangle)
result

# Area in millions of km²
abs(result$area) / 1e12
```

Multiple polygons with the `id` argument:

```{r polygon-multiple}
pts <- cbind(
  lon = c(0, 1, 1, 10, 11, 11),
  lat = c(0, 0, 1, 0, 0, 1)
)
polygon_area(pts, id = c(1, 1, 1, 2, 2, 2))
```

## Polar Regions

All functions handle polar regions automatically using UPS (Universal Polar 
Stereographic) when appropriate:

```{r polar}
polar <- cbind(c(0, 180), c(89, -89))

# MGRS uses polar grid zones
mgrs_fwd(polar)

# UTM/UPS shows zone 0 for polar
utmups_fwd(polar)
```

## Performance

All functions are implemented in C++ and fully vectorized:

```{r performance, eval=TRUE}
# 40,000+ points in milliseconds
x <- cbind(runif(40000, -180, 180), runif(40000, -80, 80))

system.time(mgrs_fwd(x))

system.time(geohash_fwd(x))

system.time(geodesic_distance_matrix(x[1:100,]))

```

## Further Reading

- [GeographicLib documentation](https://geographiclib.sourceforge.io/)
- [MGRS on Wikipedia](https://en.wikipedia.org/wiki/Military_Grid_Reference_System)
- [Geohash on Wikipedia](https://en.wikipedia.org/wiki/Geohash)
- [Geodesics on an ellipsoid](https://en.wikipedia.org/wiki/Geodesics_on_an_ellipsoid)
