---
title: "Map Projections: UTM/UPS, LCC, Cassini, Gnomonic, and Azimuthal Equidistant"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map Projections: UTM/UPS, LCC, Cassini, Gnomonic, and Azimuthal Equidistant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(geographiclib)
```

This vignette covers the map projections available in geographiclib. Each
projection has different properties making it suitable for different purposes.

## Example Locations

We'll use a mix of Northern and Southern Hemisphere locations:

```{r locations}
# Global test points
world_pts <- cbind(
  lon = c(151.21, 147.32, 166.67, -68.30, -0.13, -74.01, 139.69, 0),
  lat = c(-33.87, -42.88, -77.85, -54.80, 51.51, 40.71, 35.69, -90)
)
rownames(world_pts) <- c("Sydney", "Hobart", "McMurdo", "Ushuaia",
                          "London", "New York", "Tokyo", "South Pole")

# Australian cities for regional examples
australia <- cbind(
  lon = c(151.21, 147.32, 144.96, 153.03, 115.86, 138.60),
  lat = c(-33.87, -42.88, -37.81, -27.47, -31.95, -34.93)
)
rownames(australia) <- c("Sydney", "Hobart", "Melbourne", "Brisbane", 
                          "Perth", "Adelaide")

# Antarctic stations
antarctic <- cbind(
  lon = c(166.67, 77.85, -68.13, 39.58, -60.72),
  lat = c(-77.85, -67.60, -67.57, -69.41, -62.99)
)
rownames(antarctic) <- c("McMurdo", "Davis", "Palmer", "Progress", "Orcadas")
```

## UTM/UPS - Universal Transverse Mercator

UTM divides the Earth into 60 zones, each 6° wide. For polar regions (>84°N
or <80°S), UPS (Universal Polar Stereographic) is used instead.

### Basic Conversion

```{r utm-basic}
# Convert Australian cities
utmups_fwd(australia)
```

### Understanding UTM Zones

```{r utm-zones}
# Points at different longitudes show different zones
lon_transect <- cbind(
  lon = seq(-180, 180, by = 30),
  lat = -45
)

result <- utmups_fwd(lon_transect)
data.frame(
  lon = result$lon,
  zone = result$zone,
  northp = result$northp,
  crs = result$crs
)
```

### Polar Regions (UPS)

When zone = 0, the projection is UPS rather than UTM:

```{r utm-polar}
# Antarctic stations
utmups_fwd(antarctic)

# Note: McMurdo and the polar stations use UPS (zone 0)
# Others use standard UTM zones
```

### Convergence and Scale

UTM provides convergence angle (grid vs true north) and scale factor:
 
```{r utm-properties}
result <- utmups_fwd(australia)
data.frame(
  city = rownames(australia),
  convergence = round(result$convergence, 2),
  scale = round(result$scale, 6)
)
```

### Round-trip Conversion

```{r utm-roundtrip}
fwd <- utmups_fwd(australia)
rev <- utmups_rev(fwd$x, fwd$y, fwd$zone, fwd$northp)

# Verify accuracy
max(abs(rev$lon - australia[,1]))
max(abs(rev$lat - australia[,2]))
```

## Lambert Conformal Conic (LCC)

LCC is ideal for mid-latitude regions with greater east-west extent.
It can use one standard parallel (tangent cone) or two (secant cone).

### Single Standard Parallel

```{r lcc-single}
# Project Australia using a single standard parallel at -35°
lcc_fwd(australia, lon0 = 135, stdlat = -35)
```

### Two Standard Parallels

Two standard parallels give better scale distribution across the region:

```{r lcc-two}
# Project Australia using two standard parallels
result <- lcc_fwd(australia, lon0 = 135, stdlat1 = -18, stdlat2 = -36)
result

# Scale factor is 1.0 at the standard parallels
# and slightly different elsewhere
```

### LCC for Antarctica

```{r lcc-antarctic}
# Antarctic projection centered on the pole
lcc_fwd(antarctic, lon0 = 0, stdlat1 = -71, stdlat2 = -89)
```

### Round-trip Accuracy

```{r lcc-roundtrip}
fwd <- lcc_fwd(australia, lon0 = 135, stdlat1 = -18, stdlat2 = -36)
rev <- lcc_rev(fwd$x, fwd$y, lon0 = 135, stdlat1 = -18, stdlat2 = -36)

max(abs(rev$lon - australia[,1]))
max(abs(rev$lat - australia[,2]))
```

## Azimuthal Equidistant

This projection preserves distances from the center point. Useful for showing
distances from a specific location.

### Distances from Sydney

```{r azeq-sydney}
# Project world cities relative to Sydney
sydney <- c(151.21, -33.87)
result <- azeq_fwd(world_pts, lon0 = sydney[1], lat0 = sydney[2])
result

# Distance from Sydney (in km) = sqrt(x² + y²) / 1000
distances <- sqrt(result$x^2 + result$y^2) / 1000
data.frame(
  city = rownames(world_pts),
  distance_km = round(distances)
)
```

### Distances from South Pole

```{r azeq-pole}
# Distance from South Pole to Antarctic stations
result <- azeq_fwd(antarctic, lon0 = 0, lat0 = -90)

distances <- sqrt(result$x^2 + result$y^2) / 1000
data.frame(
  station = rownames(antarctic),
  lat = antarctic[,2],
  distance_from_pole_km = round(distances)
)
```

### Round-trip Conversion

```{r azeq-roundtrip}
fwd <- azeq_fwd(australia, lon0 = 135, lat0 = -25)
rev <- azeq_rev(fwd$x, fwd$y, lon0 = 135, lat0 = -25)

max(abs(rev$lon - australia[,1]))
max(abs(rev$lat - australia[,2]))
```

## Cassini-Soldner Projection

A historical projection used for large-scale topographic mapping. It's a
transverse cylindrical projection that preserves scale along the central
meridian.

### Regional Mapping

```{r cassini-basic}
# Tasmania centered on Hobart
tasmania <- cbind(
  lon = c(147.32, 145.49, 146.82, 148.29, 147.13),
  lat = c(-42.88, -40.83, -41.44, -42.15, -43.21)
)
rownames(tasmania) <- c("Hobart", "Launceston", "Devonport", "St Helens", "Dover")

cassini_fwd(tasmania, lon0 = 147, lat0 = -42)
```

### Cassini for Antarctic Survey

```{r cassini-antarctic}
# McMurdo area survey
mcmurdo_area <- cbind(
  lon = c(166.67, 166.40, 167.00, 166.87, 168.40),
  lat = c(-77.85, -77.55, -78.15, -78.65, -77.18)
)
rownames(mcmurdo_area) <- c("McMurdo", "Marble Point", "Black Island", 
                            "Minna Bluff", "Cape Adare")

cassini_fwd(mcmurdo_area, lon0 = 166.67, lat0 = -77.85)
```

## Gnomonic Projection

The gnomonic projection has a unique property: **geodesics (great circles)
appear as straight lines**. This makes it invaluable for route planning.

### Great Circle Routes Appear Straight

```{r gnomonic-routes}
# Project Sydney-London great circle path
sydney_london <- geodesic_path(c(151.21, -33.87), c(-0.13, 51.51), n = 10)

# Project onto gnomonic centered between them
gnomonic_fwd(cbind(sydney_london$lon, sydney_london$lat), 
             lon0 = 75, lat0 = 10)
```

### Route Planning

```{r gnomonic-planning}
# Flights from Sydney - project candidate destinations
destinations <- cbind(
  lon = c(-0.13, -74.01, 139.69, 77.22, -43.17),
  lat = c(51.51, 40.71, 35.69, 28.61, -22.91)
)
rownames(destinations) <- c("London", "New York", "Tokyo", "Delhi", "Rio")

# Gnomonic from Sydney shows great circle routes as straight lines
gnomonic_fwd(destinations, lon0 = 151.21, lat0 = -33.87)
```

### Antarctic Flight Planning

```{r gnomonic-antarctic}
# Plan routes from McMurdo
gnomonic_fwd(antarctic, lon0 = 166.67, lat0 = -77.85)

# The x,y coordinates show direction and relative great-circle distance
# (though scale varies significantly with distance from center)
```

## OSGB - Ordnance Survey National Grid

OSGB is specific to Great Britain. **Note:** It uses the OSGB36 datum, not WGS84.

```{r osgb}
# British locations (using approximate OSGB36 coordinates)
britain <- cbind(
  lon = c(-0.127, -3.188, -4.251, -1.890, -2.587),
  lat = c(51.507, 55.953, 55.864, 52.486, 51.454)
)
rownames(britain) <- c("London", "Edinburgh", "Glasgow", "Birmingham", "Cardiff")

# Convert to OSGB grid
osgb_fwd(britain)
```

### Grid References

```{r osgb-gridref}
# Get alphanumeric grid references
osgb_gridref(britain, precision = 3)  # 100m precision

# Parse a grid reference
osgb_gridref_rev("TQ308080")
```

## Projection Comparison

Different projections preserve different properties:

| Projection | Preserves | Best For |
|------------|-----------|----------|
| UTM/UPS | Shape (conformal) | Global standard, topographic maps |
| LCC | Shape (conformal) | Mid-latitude regional maps |
| Azimuthal Equidistant | Distance from center | Showing distances from a point |
| Cassini-Soldner | Scale on central meridian | Large-scale surveys |
| Gnomonic | Great circles as straight lines | Route planning |
| OSGB | Shape (conformal) | British mapping |

### Visual Comparison: Same Region, Different Projections

```{r comparison}
# Project Australia with different methods
australia_pts <- cbind(lon = c(151, 147, 145, 153, 116, 139),
                       lat = c(-34, -43, -38, -27, -32, -35))

list(
  UTM = utmups_fwd(australia_pts)[c("x", "y")],
  LCC = lcc_fwd(australia_pts, lon0 = 135, stdlat1 = -18, stdlat2 = -36)[c("x", "y")],
  AzEq = azeq_fwd(australia_pts, lon0 = 135, lat0 = -25)[c("x", "y")],
  Cassini = cassini_fwd(australia_pts, lon0 = 135, lat0 = -25)[c("x", "y")]
)
```

## See Also

- `vignette("geodesics")` for distance and bearing calculations
- `vignette("grid-reference-systems")` for MGRS, Geohash, etc.
- `vignette("local-coordinates")` for Local Cartesian (ENU) and Geocentric
- [UTM on Wikipedia](https://en.wikipedia.org/wiki/Universal_Transverse_Mercator_coordinate_system)
- [Lambert Conformal Conic](https://en.wikipedia.org/wiki/Lambert_conformal_conic_projection)
