---
title: "Three numbers, one polygon"
output: rmarkdown::github_document
---

```{r notes, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
## if this ever gets updated do this, copy the .md to https://github.com/hypertidy/geographiclib/wiki/Measurement:-three-numbers-from-one-polygon
##rmarkdown::render("vignettes/measurement.Rmd")
library(sf)
library(terra)
sf::sf_use_s2(FALSE)
```

Here is a small polygon in EPSG:2154 (RGF93 / Lambert-93, France). We want its perimeter and area. How many different answers can we get?

```{r setup}
options(digits = 9)
sf::sf_use_s2(FALSE)
## Spherical geometry (s2) switched off
xy <- cbind(
  x = c(320102.67, 320111.85, 320131.82, 320134.33, 320138.50,
        320141.93, 320135.33, 320118.58, 320113.45, 320105.18, 320102.67),
  y = c(6845954.56, 6845962.15, 6845971.55, 6845971.90, 6845953.21,
        6845936.12, 6845935.07, 6845931.54, 6845929.91, 6845950.57, 6845954.56)
)
p   <- sf::st_sf(geometry = sf::st_sfc(sf::st_polygon(list(xy)), crs = "EPSG:2154"))
pna <- p; sf::st_crs(pna) <- NA
pll <- sf::st_transform(p, "EPSG:4326")
ll  <- terra::project(xy, from = terra::crs(terra::vect(p)), to = "EPSG:4326")
```

Four scenarios: projected CRS, NA CRS, longlat with s2 off, longlat with s2 on.
geographiclib is the independent Karney reference; it operates on longlat coordinates
so the projected and NA CRS rows are not applicable.

## Scenario 1: projected CRS (EPSG:2154)

```{r scenario1}
## perimeter
as.numeric(sf::st_perimeter(p))
as.numeric(sf::st_length(sf::st_boundary(p)))
terra::perim(terra::vect(p))

## area
as.numeric(sf::st_area(p))
terra::expanse(terra::vect(p), unit = "m")

## geographiclib reference
geographiclib::polygon_area(ll)
```

sf and terra agree on perimeter (planar). For area, terra reprojects internally
and returns the ellipsoidal answer; sf returns the planar shoelace result.
geographiclib confirms the ellipsoidal value.

## Scenario 2: NA CRS

```{r scenario2}
## perimeter
as.numeric(sf::st_perimeter(pna))
terra::perim(terra::vect(pna))

## area
as.numeric(sf::st_area(pna))
terra::expanse(terra::vect(pna), unit = "m")
```

Both return planar results. terra warns ("unknown CRS, results can be wrong");
sf is silent.

## Scenario 3: longlat, s2 off

```{r scenario3}
sf::sf_use_s2(FALSE)

## perimeter
as.numeric(sf::st_perimeter(pll))
as.numeric(sf::st_length(sf::st_boundary(pll)))
terra::perim(terra::vect(pll))

## area
as.numeric(sf::st_area(pll))
terra::expanse(terra::vect(pll), unit = "m")

## geographiclib reference
geographiclib::polygon_area(ll)
```

All tools agree on the ellipsoidal (Karney) answer.

## Scenario 4: longlat, s2 on

```{r scenario4}
sf::sf_use_s2(TRUE)

## perimeter
as.numeric(sf::st_perimeter(pll))
as.numeric(sf::st_length(sf::st_boundary(pll)))
terra::perim(terra::vect(pll))

## area
as.numeric(sf::st_area(pll))
terra::expanse(terra::vect(pll), unit = "m")

## geographiclib reference (unaffected by sf_use_s2)
geographiclib::polygon_area(ll)

sf::sf_use_s2(FALSE)  ## restore
```

sf gives the spherical answer (s2 uses radius 6,371,010 m, zero flattening).
terra and geographiclib are unaffected by `sf_use_s2()` and return the
ellipsoidal answer throughout.

## Summary tables

### Perimeter (m)

| scenario | sf 1.0-21 | sf >= 1.0-22 | terra | geographiclib |
|---|---:|---:|---:|---:|
| projected (EPSG:2154) | 129.248935 | 129.248935 | 129.248935 † | — |
| NA crs | 129.248935 | 129.248935 | 129.248935 ‡ | — |
| longlat, s2 off | 129.285384 | 129.285384 | 129.285384 | 129.285384 |
| longlat, s2 on | ★129.09815  | 129.285384 | 129.285384 | 129.285384 |

### Area (m²)

| scenario | sf 1.0-21 | sf >= 1.0-22 | terra | geographiclib |
|---|---:|---:|---:|---:|
| projected (EPSG:2154) | 1049.651 | 1049.651 | 1050.243 † | — |
| NA crs | 1049.651 | 1049.651 | 1049.651 ‡ | — |
| longlat, s2 off | 1050.243 | 1050.243 | 1050.243 | 1050.243 |
| longlat, s2 on | 1047.026 | 1047.026 | 1050.243 | 1050.243 |

† `terra::perim` and `terra::expanse` are inconsistent for projected CRS: `expanse`
reprojects internally and returns the ellipsoidal answer; `perim` trusts the
projected coordinates and returns planar. No warning.

‡ terra warns on NA CRS; sf is silent.

★ fixed in sf 1.0-22 (#2541): `st_perimeter()` now uses ellipsoidal computation
when `sf_use_s2()` is FALSE.

## The inconsistencies

**1. terra::expanse vs terra::perim (projected CRS)**
`expanse` reprojects to longlat internally and applies Karney for any known CRS.
`perim` does not — it returns planar distance for projected inputs and only switches
to Karney for longlat. This asymmetry presents a cryptic potential gotch for users; `expanse` has the more defensible design as terra does generally. 

**2. sf projected CRS: no ellipsoidal awareness for area**
`sf::st_area()` on a projected polygon returns the planar shoelace result with no
warning. It does not reproject. The answer is "correct in the map" but not on
the Earth. `terra::expanse` does the extra work silently.

**3. sf NA CRS: silent vs terra's warning**
Both return planar results with no CRS. terra warns; sf does not.

**4. sf < 1.0-22: st_perimeter ignored sf_use_s2()**
With longlat input and `sf_use_s2(FALSE)`, `st_perimeter()` still dispatched
through s2 giving the spherical answer, while `st_length(st_boundary())` correctly
gave Karney. Fixed in sf 1.0-22 (#2541). 

## What to use

For consistent ellipsoidal results across all CRS inputs: `terra::expanse()` for
area, `geodist::geodist(..., measure = "geodesic")` or `geographiclib` for
perimeter/length. Note that `terra::perim()` requires longlat input to give the
ellipsoidal answer.

If you use sf: transform to longlat and set `sf_use_s2(FALSE)`. Both `st_area()`
and `st_length(st_boundary())` then give Karney results. Keep in mind this is a
global option — set it deliberately and restore it.

There are other functions in the families of sf and terra for distance calculations (i.e. see package {lwgeom} and {geosphere}) but this article is motivated by the problematic framing users face around expectations and results with very widely used functions. 

If you want truly xy-in-map only calculations the only really reliable way is to set crs to NA, this was established by sf early in its history as "good practice" and unfortunately is a clear conclusion from the behaviour we see  here. Having functions that do clearly what they seem named to do would be preferable, but as it is users have to navigate these vagaries to be sure of what they are getting. These same principles appy to buffering and nearest-calcuations but we will not explore those implications here besides to point out one example: 

```r
Warning message:
In st_voronoi.sfc(st_geometry(x), st_sfc(envelope), dTolerance,  :
  st_voronoi does not correctly triangulate longitude/latitude data
```

When seeing warnings like this you can wonder "what does correctly" mean here? As we've seen
there is more than one version of "correct" because that depends on what you intend to do 
and purpose of your task. A function should simply do what it advertises and (ideally) be consistent within its own context. Declarations of what is correct without significantly more context are specious, at best. 

## Versions

```{r session}
R.version.string

packageVersion("geographiclib")
packageVersion("terra")
packageVersion("sf")

```

---

*Code from these gists:*

- [correct_area.md](https://gist.github.com/mdsumner/6caeefaeb5200ddf5e77d2aeced3bebd) — area comparison across projections and engines (2023)
- [area_and_crs.md](https://gist.github.com/mdsumner/5c52e770d822bc4ac183e7460cb2b4fc) — area with Mercator distortion (2024)
- [real_dist.md](https://gist.github.com/mdsumner/5daedfcbee851ca31a7aa9006dbf74cf) — perimeter, three numbers, R-sig-geo May 2025
