---
title: "Three numbers, one polygon"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Three numbers, one polygon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Here is a small polygon in EPSG:2154 (RGF93 / Lambert-93, France). We want its perimeter and area. How many different answers can we get?

```{r setup}
options(digits = 9)
sf::sf_use_s2(FALSE)

xy <- cbind(
  x = c(320102.67, 320111.85, 320131.82, 320134.33, 320138.50,
        320141.93, 320135.33, 320118.58, 320113.45, 320105.18, 320102.67),
  y = c(6845954.56, 6845962.15, 6845971.55, 6845971.90, 6845953.21,
        6845936.12, 6845935.07, 6845931.54, 6845929.91, 6845950.57, 6845954.56)
)
p   <- sf::st_sf(geometry = sf::st_sfc(sf::st_polygon(list(xy)), crs = "EPSG:2154"))
pna <- p; sf::st_crs(pna) <- NA
pll <- sf::st_transform(p, "EPSG:4326")
ll  <- terra::project(xy, from = terra::crs(terra::vect(p)), to = "EPSG:4326")
```

Four scenarios: projected CRS, NA CRS, longlat with s2 off, longlat with s2 on.
geographiclib is the independent Karney reference; it operates on longlat coordinates
so the projected and NA CRS rows are not applicable.

## Scenario 1: projected CRS (EPSG:2154)

```{r scenario1}
## perimeter
as.numeric(sf::st_perimeter(p))
as.numeric(sf::st_length(sf::st_boundary(p)))
terra::perim(terra::vect(p))

## area
as.numeric(sf::st_area(p))
terra::expanse(terra::vect(p), unit = "m")

## geographiclib reference
geographiclib::polygon_area(ll)
```

sf and terra agree on perimeter (planar). For area, terra reprojects internally
and returns the ellipsoidal answer; sf returns the planar shoelace result.
geographiclib confirms the ellipsoidal value.

## Scenario 2: NA CRS

```{r scenario2}
## perimeter
as.numeric(sf::st_perimeter(pna))
terra::perim(terra::vect(pna))

## area
as.numeric(sf::st_area(pna))
terra::expanse(terra::vect(pna), unit = "m")
```

Both return planar results. terra warns ("unknown CRS, results can be wrong");
sf is silent.

## Scenario 3: longlat, s2 off

```{r scenario3}
sf::sf_use_s2(FALSE)

## perimeter
as.numeric(sf::st_perimeter(pll))
as.numeric(sf::st_length(sf::st_boundary(pll)))
terra::perim(terra::vect(pll))

## area
as.numeric(sf::st_area(pll))
terra::expanse(terra::vect(pll), unit = "m")

## geographiclib reference
geographiclib::polygon_area(ll)
```

All tools agree on the ellipsoidal (Karney) answer.

## Scenario 4: longlat, s2 on

```{r scenario4}
sf::sf_use_s2(TRUE)

## perimeter
as.numeric(sf::st_perimeter(pll))
as.numeric(sf::st_length(sf::st_boundary(pll)))
terra::perim(terra::vect(pll))

## area
as.numeric(sf::st_area(pll))
terra::expanse(terra::vect(pll), unit = "m")

## geographiclib reference (unaffected by sf_use_s2)
geographiclib::polygon_area(ll)

sf::sf_use_s2(FALSE)  ## restore
```

sf gives the spherical answer (s2 uses radius 6,371,010 m, zero flattening).
terra and geographiclib are unaffected by `sf_use_s2()` and return the
ellipsoidal answer throughout.

## Summary tables

### Perimeter (m)

| scenario | sf 1.0-21 | sf >= 1.0-22 | terra | geographiclib |
|---|---:|---:|---:|---:|
| projected (EPSG:2154) | 129.248935 | 129.248935 | 129.248935 † | — |
| NA crs | 129.248935 | 129.248935 | 129.248935 ‡ | — |
| longlat, s2 off | 129.285384 | 129.285384 | 129.285384 | 129.285384 |
| longlat, s2 on | 129.09815 ★ | 129.285384 | 129.285384 | 129.285384 |

### Area (m²)

| scenario | sf 1.0-21 | sf >= 1.0-22 | terra | geographiclib |
|---|---:|---:|---:|---:|
| projected (EPSG:2154) | 1049.651 | 1049.651 | 1050.243 † | — |
| NA crs | 1049.651 | 1049.651 | 1049.651 ‡ | — |
| longlat, s2 off | 1050.243 | 1050.243 | 1050.243 | 1050.243 |
| longlat, s2 on | 1047.026 | 1047.026 | 1050.243 | 1050.243 |

† `terra::perim` and `terra::expanse` are inconsistent for projected CRS: `expanse`
reprojects internally and returns the ellipsoidal answer; `perim` trusts the
projected coordinates and returns planar. No warning.

‡ terra warns on NA CRS; sf is silent.

★ fixed in sf 1.0-22 (#2541): `st_perimeter()` now uses ellipsoidal computation
when `sf_use_s2()` is FALSE.

## The inconsistencies

**1. terra::expanse vs terra::perim (projected CRS)**
`expanse` reprojects to longlat internally and applies Karney for any known CRS.
`perim` does not — it returns planar distance for projected inputs and only switches
to Karney for longlat. There is no good reason for this asymmetry; `expanse` has
the more defensible design. Worth a terra issue.

**2. sf projected CRS: no ellipsoidal awareness for area**
`sf::st_area()` on a projected polygon returns the planar shoelace result with no
warning. It does not reproject. The answer is "correct in the map" but not on
the Earth. `terra::expanse` does the extra work silently.

**3. sf NA CRS: silent vs terra's warning**
Both return planar results with no CRS. terra warns; sf does not.

**4. sf < 1.0-22: st_perimeter ignored sf_use_s2()**
With longlat input and `sf_use_s2(FALSE)`, `st_perimeter()` still dispatched
through s2 giving the spherical answer, while `st_length(st_boundary())` correctly
gave Karney. Fixed in sf 1.0-22 (#2541). The broader issue — that s2-on silently
gives spherical rather than ellipsoidal — remains.

## What to use

For consistent ellipsoidal results across all CRS inputs: `terra::expanse()` for
area, `geodist::geodist(..., measure = "geodesic")` or `geographiclib` for
perimeter/length. Note that `terra::perim()` requires longlat input to give the
ellipsoidal answer.

If you use sf: transform to longlat and set `sf_use_s2(FALSE)`. Both `st_area()`
and `st_length(st_boundary())` then give Karney results. Keep in mind this is a
global option — set it deliberately and restore it.

## Session info

```{r session}
sessioninfo::session_info()
```

---

*Code from these gists:*

- [correct_area.md](https://gist.github.com/mdsumner/6caeefaeb5200ddf5e77d2aeced3bebd) — area comparison across projections and engines (2023)
- [area_and_crs.md](https://gist.github.com/mdsumner/5c52e770d822bc4ac183e7460cb2b4fc) — area with Mercator distortion (2024)
- [real_dist.md](https://gist.github.com/mdsumner/5daedfcbee851ca31a7aa9006dbf74cf) — perimeter, three numbers, R-sig-geo May 2025
